---
title: "Cybersecurity Threat Analysis Dashboard"
author: "Zhiming Zhang @zzknight2016@gmail.com"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: embed
    navbar:
      - { title: "Overview", href: "#overview", align: left }
      - { title: "Trends", href: "#trends", align: left }
      - { title: "Networks", href: "#networks", align: left }
      - { title: "Clustering", href: "#clustering", align: left }
---

```{r setup, include=FALSE}
library(flexdashboard)
library(highcharter)
library(dplyr)
library(viridisLite)
library(forecast)
library(treemap)
library(arules)
library(readr)
library(ggplot2)
library(corrplot)
library(zoo)
library(tidyr)
library(networkD3)
library(ggrepel)
library(ggraph)
library(visNetwork)
library(bnlearn)
library(igraph)

# Custom theme
thm <- 
  hc_theme(
    colors = c("#1a6ecc", "#434348", "#90ed7d", "#f7a35c", "#8085e9"),
    chart = list(
      backgroundColor = "transparent",
      style = list(fontFamily = "Source Sans Pro")
    ),
    xAxis = list(
      gridLineWidth = 1
    )
  )

```


```{r data_loading_preprocessing, include=FALSE}
# Data Loading
df1209 <- read_csv("Data/2022-12-09-enriched.csv")
df0704 <- read_csv("Data/2022-07-04-enriched.csv")
df0627 <- read_csv("Data/2022-06-27-enriched.csv")
df0609 <- read_csv("Data/2022-06-09-enriched.csv")
df0608 <- read_csv("Data/2022-06-08-enriched.csv")

all_df <- rbind(df0608, df0609, df0627, df0704, df1209)

# remove "notes" column which contains no valuable info
all_df <- subset(all_df, select = -c(notes))

# Impute missing values using the first non-missing value with the same cve_id
# (see the R script for more information)
all_imp <- all_df %>%
  group_by(cve_id) %>%
  mutate(across(everything(), ~if_else(is.na(.), first(.[!is.na(.)]), .)))

# Impute product and short description columns
all_imp$product[is.na(all_imp$product)] <- "fuel cms"
all_imp$short_description[is.na(all_imp$short_description)] <- "na"

# Impute pub_date for NA values based on cve_id
all_imp <- all_imp %>%
  group_by(cve_id) %>%
  mutate(pub_date = ifelse(
    is.na(pub_date), 
    if(all(is.na(pub_date))) NA else max(pub_date, na.rm = TRUE), 
    pub_date
  ))

all_imp <- na.omit(all_imp)

# Create numerical version for analysis
all_num_clean <- all_imp
all_num_clean$product <- as.numeric(as.factor(all_imp$product))
all_num_clean$vulnerability_name <- as.numeric(as.factor(all_imp$vulnerability_name))
all_num_clean$short_description <- as.numeric(as.factor(all_imp$short_description))
all_num_clean$required_action <- as.numeric(as.factor(all_imp$required_action))
all_num_clean$cwe <- as.numeric(as.factor(all_imp$cwe))
all_num_clean$vector <- as.numeric(as.factor(all_imp$vector))
all_num_clean$complexity <- as.numeric(as.factor(all_imp$complexity))
all_num_clean$severity <- as.numeric(as.factor(all_imp$severity))
all_num_clean$vendor_project <- as.numeric(as.factor(all_imp$vendor_project))
all_num_clean$cve_id <- as.numeric(as.factor(all_imp$cve_id))
all_num_clean$date_added <- as.numeric(as.factor(all_imp$date_added))
all_num_clean$due_date <- as.numeric(as.factor(all_imp$due_date))
all_num_clean$pub_date <- as.numeric(as.factor(all_imp$pub_date))

# Calculate patching time
all_imp$patching_time <- as.numeric(all_imp$due_date - all_imp$pub_date)
```

Overview {data-navmenu="Analysis"}
=====================================

Column {data-width=400}
-----------------------------------------------------------------------

### Dataset Summary

```{r}
valueBox(
  value = nrow(all_imp),
  caption = "Total Vulnerabilities",
  icon = "fa-bug",
  color = "danger"
)
```

### Unique CVEs

```{r}
valueBox(
  value = length(unique(all_imp$cve_id)),
  caption = "Unique CVE IDs",
  icon = "fa-shield-alt",
  color = "warning"
)
```

### Vendors Affected

```{r}
valueBox(
  value = length(unique(all_imp$vendor_project)),
  caption = "Unique Vendors",
  icon = "fa-building",
  color = "info"
)
```

### Products Affected

```{r}
valueBox(
  value = length(unique(all_imp$product)),
  caption = "Unique Products",
  icon = "fa-cogs",
  color = "success"
)
```

Column {data-width=600}
-----------------------------------------------------------------------

### Vulnerability Severity Distribution

```{r}
severity_counts <- table(all_imp$severity)
severity_df <- data.frame(
  severity = names(severity_counts),
  count = as.numeric(severity_counts)
)

hchart(severity_df, "pie", hcaes(x = severity, y = count), name = "Vulnerabilities") %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Distribution of Vulnerability Severity Levels") %>%
  hc_tooltip(pointFormat = "<b>{point.name}</b>: {point.y} ({point.percentage:.1f}%)")
```

### Top 10 Vendors by Vulnerability Count

```{r}
top_vendors <- all_imp %>%
  group_by(vendor_project) %>%
  summarize(total_vulnerabilities = n()) %>%
  arrange(desc(total_vulnerabilities)) %>%
  head(10)

hchart(top_vendors, "column", hcaes(x = vendor_project, y = total_vulnerabilities)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Top 10 Vendors with Highest Number of Vulnerabilities") %>%
  hc_xAxis(title = list(text = "Vendor")) %>%
  hc_yAxis(title = list(text = "Total Vulnerabilities")) %>%
  hc_plotOptions(column = list(dataLabels = list(enabled = TRUE)))
```


### Top 10 Products with Most Vulnerabilities

```{r}
top_products <- all_imp %>%
  group_by(product) %>%
  summarize(total_vulnerabilities = n()) %>%
  arrange(desc(total_vulnerabilities)) %>%
  head(10)

hchart(top_products, "bar", hcaes(x = product, y = total_vulnerabilities)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Top 10 Products by Vulnerability Count") %>%
  hc_xAxis(title = list(text = "Product")) %>%
  hc_yAxis(title = list(text = "Total Vulnerabilities"))
```


Column {data-width=600}
-----------------------------------------------------------------------

### Vulnerability by Vector Type

```{r}
vector_counts <- all_imp %>%
  group_by(vector) %>%
  summarise(count = n()) %>%
  drop_na()

hchart(vector_counts, "bar", hcaes(x = vector, y = count)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Distribution of Vulnerabilities by Attack Vector") %>%
  hc_xAxis(title = list(text = "Attack Vector")) %>%
  hc_yAxis(title = list(text = "Count"))
```

### Complexity Level Distribution

```{r}
complexity_counts <- table(all_imp$complexity)
complexity_df <- data.frame(
  complexity = names(complexity_counts),
  count = as.numeric(complexity_counts)
)

hchart(complexity_df, "column", hcaes(x = complexity, y = count)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Vulnerability Distribution by Complexity Level") %>%
  hc_xAxis(title = list(text = "Complexity Level")) %>%
  hc_yAxis(title = list(text = "Count")) %>%
  hc_plotOptions(column = list(dataLabels = list(enabled = TRUE)))
```


### Most Common CWE Categories

```{r}
common_cwe <- all_imp %>%
  group_by(cwe) %>%
  summarize(total_vulnerabilities = n()) %>%
  arrange(desc(total_vulnerabilities)) %>%
  head(10)

hchart(common_cwe, "column", hcaes(x = cwe, y = total_vulnerabilities)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Most Common CWE Categories") %>%
  hc_xAxis(title = list(text = "CWE Category")) %>%
  hc_yAxis(title = list(text = "Count")) %>%
  hc_plotOptions(column = list(dataLabels = list(enabled = TRUE)))
```


Patching Speed {data-navmenu="Analysis"}
=====================================

Column {data-width=600}
-----------------------------------------------------------------------
###  Patching Speed Distribution
```{r}
# Calculate patching time
all_imp$patching_time <- all_imp$due_date - all_imp$pub_date
all_imp$patching_time <- as.numeric(all_imp$patching_time)

# Create histogram data for highcharter
hist_data <- hist(all_imp$patching_time, breaks = 20, plot = FALSE)
hist_df <- data.frame(
  x = hist_data$mids,
  y = hist_data$counts
)

# Create interactive histogram with highcharter
hchart(hist_df, "column", hcaes(x = x, y = y)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Distribution of Patching Time") %>%
  hc_xAxis(title = list(text = "Patching Time (Days)")) %>%
  hc_yAxis(title = list(text = "Frequency")) %>%
  hc_tooltip(
    pointFormat = "Days: {point.x:.1f}<br/>Count: {point.y}"
  ) %>%
  hc_plotOptions(
    column = list(
      borderWidth = 0,
      groupPadding = 0,
      pointPadding = 0.05
    )
  )
```

### Patching Speed Analysis

```{r}
# Calculate average patching time by vendor
patching_speed <- all_imp %>%
  filter(!is.na(patching_time)) %>%
  group_by(vendor_project) %>%
  summarise(avg_patching_time = mean(patching_time, na.rm = TRUE),
            count = n()) %>%
  filter(count >= 5) %>%  # Only vendors with at least 5 vulnerabilities
  arrange(avg_patching_time) %>%
  head(15)

hchart(patching_speed, "column", hcaes(x = vendor_project, y = avg_patching_time)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Average Patching Time by Vendor (days)") %>%
  hc_xAxis(title = list(text = "Vendor")) %>%
  hc_yAxis(title = list(text = "Average Days to Patch")) %>%
  hc_plotOptions(column = list(dataLabels = list(enabled = TRUE, format = "{point.y:.0f}")))
```


Trends {data-navmenu="Analysis"}
=====================================

Column {data-width=600}
-----------------------------------------------------------------------

### Summary

```{r}
```

Column {data-width=600}
-----------------------------------------------------------------------

### Vulnerability Trends Over Time

```{r}
# Convert date_added column to year-month format
all_imp$date_added_ym <- as.yearmon(all_imp$date_added)
vuln_counts <- table(all_imp$date_added_ym)

# Create trend data frame with proper date conversion
trend_df <- data.frame(
  date_ym = names(vuln_counts),
  count = as.numeric(vuln_counts)
) %>%
  mutate(
    date = as.Date(as.yearmon(date_ym))
  )

hchart(trend_df, "line", hcaes(x = date, y = count)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Vulnerability Discovery Trends Over Time") %>%
  hc_xAxis(title = list(text = "Date"), type = "datetime") %>%
  hc_yAxis(title = list(text = "Number of Vulnerabilities")) %>%
  hc_tooltip(xDateFormat = "%Y-%m")
```

### Severity Trends by Time

```{r}
severity_time <- all_imp %>%
  mutate(date_added_ym = as.yearmon(date_added)) %>%
  group_by(date_added_ym, severity) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(date = as.Date(date_added_ym))

# Handle NA values and filter valid data
severity_time <- severity_time %>%
  filter(!is.na(date) & !is.na(severity))

severity_colors <- c("CRITICAL" = "#d32f2f", "HIGH" = "#f57c00", "MEDIUM" = "#fbc02d", "LOW" = "#388e3c")

if(nrow(severity_time) > 0) {
  hchart(severity_time, "line", hcaes(x = date, y = count, group = severity)) %>%
    hc_add_theme(thm) %>%
    hc_title(text = "Vulnerability Trends by Severity Over Time") %>%
    hc_xAxis(title = list(text = "Date"), type = "datetime") %>%
    hc_yAxis(title = list(text = "Count")) %>%
    hc_colors(unname(severity_colors[names(severity_colors) %in% unique(severity_time$severity)]))
} else {
  # Fallback static chart if data is problematic
  severity_counts <- table(all_imp$severity)
  severity_df <- data.frame(
    severity = names(severity_counts),
    count = as.numeric(severity_counts)
  )
  hchart(severity_df, "column", hcaes(x = severity, y = count)) %>%
    hc_add_theme(thm) %>%
    hc_title(text = "Vulnerability Count by Severity")
}
```


Networks {data-navmenu="Analysis"}
=====================================

Column {data-width=600}
-----------------------------------------------------------------------

### Vendor-Product Network
```{r}
# Create a data frame of nodes
nodes <- data.frame(id = unique(c(all_imp$vendor_project, all_imp$product)), 
                    label = unique(c(all_imp$vendor_project, all_imp$product)))

# Create a data frame of edges
edges <- data.frame(from = all_imp$vendor_project, to = all_imp$product)

# Create the visNetwork object with nodes and edges data
network <- visNetwork(nodes, edges) %>%
  visOptions(width = "100%", height = "500px") %>%
  visInteraction(hover = TRUE, hoverConnectedEdges = TRUE) %>%
  visPhysics(enabled = TRUE)

network
```

### Network with Fruchterman-Reingold Layout 
```{r}

network_data <- all_imp[, c("vendor_project", "product")]

# Create edge list
edge_list <- as.matrix(network_data)

graph <- graph_from_edgelist(edge_list, directed = TRUE)

# Create layout using Fruchterman-Reingold algorithm
layout <- layout_with_fr(graph)

# Convert graph layout to data frame
layout_df <- data.frame(layout)

# Add vendor/product names to layout data frame
layout_df$names <- V(graph)$name

graph_plot <- ggraph(graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point(shape = 21, fill = "#005c87", color = "gray") +
  geom_node_text(aes(label = name), repel = TRUE, box.padding = 0.5, size = 3, max.overlaps = 20) +
  theme_void()

graph_plot
```

Column {data-width=400}
-----------------------------------------------------------------------

### Network Centrality Analysis

```{r}
# Calculate centrality measures - explicitly use igraph::degree
g <- graph_from_data_frame(all_imp, directed = FALSE)
centrality <- igraph::degree(g)  # Explicitly use igraph::degree
top_central <- head(sort(centrality, decreasing = TRUE), 10)


central_df <- data.frame(
  entity = names(top_central),
  centrality = as.numeric(top_central)
)

hchart(central_df, "bar", hcaes(x = entity, y = centrality)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Top 10 Most Central Entities in Network") %>%
  hc_xAxis(title = list(text = "Entity")) %>%
  hc_yAxis(title = list(text = "Degree Centrality"))
```


Clustering {data-navmenu="Analysis"}
=====================================

Column {data-width=600}
-----------------------------------------------------------------------

### CWE vs Severity Heatmap

```{r}
# Create heatmap data
heatmap_data <- all_num_clean %>%
  group_by(cwe, severity) %>%
  summarise(count = n(), .groups = 'drop') %>%
  filter(count > 0)

# Convert to matrix format for heatmap
hchart(heatmap_data, "heatmap", hcaes(x = cwe, y = severity, value = count)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Vulnerability Clusters: CWE vs Severity") %>%
  hc_xAxis(title = list(text = "CWE Category")) %>%
  hc_yAxis(title = list(text = "Severity Level")) %>%
  hc_colorAxis(min = 0, minColor = "#FFFFFF", maxColor = "#1a6ecc") %>%
  hc_tooltip(pointFormat = "CWE: {point.x}<br>Severity: {point.y}<br>Count: <b>{point.value}</b>")
```

### Vector vs Severity Analysis

```{r}
vector_severity_data <- all_imp %>%
  group_by(vector, severity) %>%
  summarise(count = n(), .groups = 'drop') %>%
  filter(!is.na(vector) & !is.na(severity))

hchart(vector_severity_data, "column", hcaes(x = vector, y = count, group = severity)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Vulnerability Distribution: Vector vs Severity") %>%
  hc_xAxis(title = list(text = "Attack Vector")) %>%
  hc_yAxis(title = list(text = "Count")) %>%
  hc_plotOptions(column = list(stacking = "normal"))
```

Column {data-width=400}
-----------------------------------------------------------------------

### CVSS Score Distribution

```{r}
cvss_data <- all_imp %>%
  filter(!is.na(cvss)) %>%
  mutate(cvss_range = cut(cvss, breaks = c(0, 3, 6, 9, 10), 
                         labels = c("Low (0-3)", "Medium (3-6)", "High (6-9)", "Critical (9-10)")))

cvss_dist <- cvss_data %>%
  group_by(cvss_range) %>%
  summarise(count = n())

hchart(cvss_dist, "pie", hcaes(x = cvss_range, y = count)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "CVSS Score Distribution") %>%
  hc_tooltip(pointFormat = "<b>{point.name}</b>: {point.y} ({point.percentage:.1f}%)")
```

### Correlation Matrix

```{r}
# Select numeric columns for correlation
cor_data <- all_num_clean[, c("severity", "cvss", "complexity", "vector")]
cor_matrix <- cor(cor_data, use = "complete.obs")

# Convert correlation matrix to long format
cor_long <- expand.grid(Var1 = rownames(cor_matrix), Var2 = colnames(cor_matrix))
cor_long$value <- as.vector(cor_matrix)

hchart(cor_long, "heatmap", hcaes(x = Var1, y = Var2, value = value)) %>%
  hc_add_theme(thm) %>%
  hc_title(text = "Correlation Matrix of Key Variables") %>%
  hc_xAxis(title = list(text = "")) %>%
  hc_yAxis(title = list(text = "")) %>%
  hc_colorAxis(min = -1, max = 1, minColor = "#d32f2f", maxColor = "#1976d2") %>%
  hc_tooltip(pointFormat = "Correlation: <b>{point.value:.3f}</b>")
```

Summary {data-navmenu="Insights"}
=====================================

Column {data-width=600}
-----------------------------------------------------------------------

### Key Findings {.no-title}

**Dataset Overview:**
- **Total Vulnerabilities:** `r nrow(all_imp)` vulnerabilities across `r length(unique(all_imp$vendor_project))` vendors
- **Time Period:** Data spanning from 2022-06-08 to 2022-12-09
- **Coverage:** `r length(unique(all_imp$product))` unique products affected

**Severity Distribution:**
- **Critical:** `r round(sum(all_imp$severity == "CRITICAL")/nrow(all_imp)*100, 1)`% of vulnerabilities
- **High:** `r round(sum(all_imp$severity == "HIGH")/nrow(all_imp)*100, 1)`% of vulnerabilities  
- **Medium:** `r round(sum(all_imp$severity == "MEDIUM")/nrow(all_imp)*100, 1)`% of vulnerabilities
- **Low:** `r round(sum(all_imp$severity == "LOW")/nrow(all_imp)*100, 1)`% of vulnerabilities

**Top Risk Areas:**
- **Primary Attack Vector:** Network (`r round(sum(all_imp$vector == "NETWORK", na.rm=TRUE)/sum(!is.na(all_imp$vector))*100, 1)`% of vulnerabilities)
- **Most Affected Vendor:** Microsoft with `r sum(all_imp$vendor_project == "microsoft", na.rm=TRUE)` vulnerabilities
- **Most Affected Product:** Windows with `r sum(all_imp$product == "windows", na.rm=TRUE)` vulnerabilities

**Patching Performance:**
- **Average Patching Time:** `r round(mean(all_imp$patching_time, na.rm=TRUE), 0)` days
- **Median Patching Time:** `r round(median(all_imp$patching_time, na.rm=TRUE), 0)` days
- **Range:** `r min(all_imp$patching_time, na.rm=TRUE)` to `r max(all_imp$patching_time, na.rm=TRUE)` days

Column {data-width=400}
-----------------------------------------------------------------------

### Recommendations {.no-title}

**Immediate Actions:**

1. **Prioritize Critical & High Severity:** Focus on the `r round((sum(all_imp$severity == "CRITICAL") + sum(all_imp$severity == "HIGH"))/nrow(all_imp)*100, 0)`% of vulnerabilities classified as Critical or High severity

2. **Network Security:** Strengthen network-based defenses as `r round(sum(all_imp$vector == "NETWORK", na.rm=TRUE)/sum(!is.na(all_imp$vector))*100, 0)`% of vulnerabilities use network attack vectors

3. **Microsoft Products:** Implement enhanced monitoring for Microsoft products, particularly Windows, given their high vulnerability count

**Strategic Improvements:**

4. **Reduce Patching Time:** Current average of `r round(mean(all_imp$patching_time, na.rm=TRUE), 0)` days should be reduced, especially for critical vulnerabilities

5. **Vendor Engagement:** Establish dedicated security liaisons with top vendors (Microsoft, Cisco, Apple, Google)

6. **Complexity Assessment:** `r round(sum(all_imp$complexity == "LOW", na.rm=TRUE)/sum(!is.na(all_imp$complexity))*100, 0)`% of vulnerabilities are low complexity - these should be prioritized for quick wins

**Long-term Strategy:**

7. **Trend Monitoring:** Establish continuous monitoring for vulnerability trends and emerging threat patterns

8. **Network Analysis:** Use relationship mapping between vendors and products to identify potential cascade effects

9. **Predictive Analytics:** Implement machine learning models to predict vulnerability emergence and optimize resource allocation